---
title: "Lab 01 Basic of R, RStudio, and GitHub"
subtitle: "RE 519 Data Analytics and Visualization | Autumn 2025"
author: "" 
date: ""
output: 
  cleanrmd::html_document_clean:
    theme: almond
    toc: true 
    toc_depth: 3
    css: custom.css
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
    md_extensions: +tex_math_dollars+tex_math_single_backslash   
    df_print: paged    
---

<!-- Notes about the HTML and CSS file

I use the R package `cleanrmd` (https://github.com/gadenbuie/cleanrmd) which includes many beautiful themes to render this page. Also, I made some customization using the CSS file provided in the same folder. You may need install `cleanrmd` and have the CSS file in the same folder with `rmd` file to knit the same page. If you want to use the default style. 

You can change the output below to html_document, like:

title: "Lab 01 Basic of R, RStudio, and GitHub"
subtitle: "RE 519 Data Analytics and Visualization | Autumn 2025"
author: "" 
date: ""
output: html_document

--> 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

<!-- start of the main lab content --> 

**Welcome to start the journey in data science!** Through this quarter, we will 
learn how to conduct data analysis and visualization using [R](http://www.r-project.org),
[RStudio](https://posit.co/products/open-source/rstudio), and [Tableau](https://www.tableau.com). 
There are tons of resources, both online and within University of Washington, for you to start learning data science, check [resource page](https://www.yuehaoyu.com/data-analytics-visualization/resources/).

We will have **8** labs in total (2~3 classes for each lab) and you only need to submit
**once** after each lab. The due day for each lab can be found on the [course wesbite](https://www.yuehaoyu.com/data-analytics-visualization/). The submission should include Rmd, html, and any other files required to rerun the codes. 

Please check [grading policy of this class](https://www.yuehaoyu.com/data-analytics-visualization/syllabus/#course-requirements) in advance, including grade curving, late date policy, etc. **For Lab 1–3, the use of any generative AI tool (ChatGPT, Copilot, etc.) is prohibited**. We will run AI detection on each submission.
More information about [Academic Integrity](https://www.yuehaoyu.com/data-analytics-visualization/syllabus/#academic-integrity) and [the Use of AI](https://www.yuehaoyu.com/data-analytics-visualization/syllabus/#use-of-ai).

> **Recommended Readings**:  
> [Appendix B and D, Modern Data Science with R. Baumer et al. 2024.](https://mdsr-book.github.io/mdsr3e/)  
> [Chapter 2, Data Visualization. Kieran Healy. 2019.](https://staff.washington.edu/yohaoyu/data-analytics-visualization/Data-Visualization-A-Practical-Introduction.pdf)

---

## Lab 01-A: Setup

### Why R and RStudio?

R is a programming language built for statistical computing. But, if one already knows Stata or similar software, why use R? 

- R is *free*, so you don't need a terminal server.
- R has a *very* large community.
- R can handle virtually any data format.
- R makes replication easy.
- R is a *language* so it can do *everything*.
- R skills transfer to other languages like [Python](https://www.python.org) and [Julia](https://julialang.org).

R Studio is a "front-end" or integrated development environment (IDE) for R that can make your life *easier*. We'll show RStudio can...

- Organize your code, output, and plots
- Auto-complete code and highlight syntax
- Help view data and objects
- Enable easy integration of R code into documents with *R Markdown*
- Manage [`git`](https://git-scm.com) repositories
- Run interactive tutorials

### Selling you on R Markdown

The ability to create R Markdown files is a powerful advantage of R:

- Document analyses by combining text, code, and output
  - No copying and pasting into Word
  - Easy for collaborators to understand
  - Show as little or as much code as you want
- Produce many different document types as output
  - PDF documents
  - HTML webpages and reports
  - Word and PowerPoint documents 
  - Presentations 
-  Works with [$\LaTeX$](https://www.latex-project.org) and HTML for math and more formatting control

### R and RStudio Installation

Try to complete the setup before the first class. The goal of this assignment is simply to make sure that you have R up and running and can open the source document of this page, a `rmd` (R Markdown) file. 

If R & RStudio aren't running and installing properly you may need to update your operating system. Don't worry too much if you can't get them installed and running, I just want everyone to make an attempt which should save you time later on.

1. Download and install R, see here: [https://cran.r-project.org/](https://cran.r-project.org/)
2. Download and install RStudio Desktop, see here: [https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/)
3. [Windows video guide](https://www.youtube.com/watch?v=TsnGd6p9oTk) and [Mac video guide](https://www.youtube.com/watch?v=8NvvydRwxEI).

The latest version of R is **4.5.1 (Great Square Root).** You can check by typing `R.version` into your RStudio Console or run it in one chunk. 

### Getting Started

Open up RStudio now and try to download the source file, a `rmd` (R Markdown) file [here](https://www.yuehaoyu.com/data-analytics-visualization/labs/lab-01.zip) and open it on your local machine.

Use `install.packages("[package name]")` to install any required packages, or R Studio may automatically recognize that there are some packages you need to install and will install them for you.

Then, let's get oriented with the interface:

- *Top Left*: Code **editor** pane, data viewer (browse with tabs)
- *Bottom Left*: **Console** for running code (`>` prompt)
- *Top Right*: List of objects in **environment**, code **history** tab.
- *Bottom Right*: Tabs for browsing files, viewing plots, managing packages, and viewing help files.

You can change the layout and theme in *Preferences > Pane Layout/Appearance*.

---

## Lab 01-B: Basic of R/RStudio and Markdown

### R Markdown

Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>. In the lab, we will review some basic Markdown syntax: [Markdown Live Preview](https://markdownlivepreview.com).

You can put some math $y= \left(\frac{2}{3} \right)^2$ right up in there using [$\LaTeX$](https://www.latex-project.org). 
$$\frac{1}{n} \sum_{i=1}^{n} x_i = \bar{x}_n$$
*Note:* $\LaTeX$ is not required in this class. If you are curious about how mathematicians compile their books, you may refer to the [Learn LaTeX in 30 minutes](https://www.overleaf.com/learn/latex/Learn_LaTeX_in_30_minutes) from [Overleaf](https://www.overleaf.com/). Overleaf is the most popular collaborative cloud-based $\LaTeX$ editor and UW students can get [Overleaf Premium account for free](https://it.uw.edu/uware/overleaf/).

#### R Markdown Documents

This document is a R Markdown file, you can create a new one by:

1. Choose *File > New File > R Markdown...*
2. Make sure *HTML Output* is selected and click OK (select PDF if you'd like PDF as outputs)
3. Save the file somewhere, call it `any_name.Rmd`
4. Click the *Knit HTML* button
5. Watch the progress in the R Markdown pane, then gaze upon your result!

You may also open up the HTML file in your computer's browser if you so desire, using the *Open in Browser* button at the top of the preview window.

#### R Markdown Headers

The header of an .Rmd file is a [YAML](http://yaml.org/) code block, and everything else is part of the main document. To mess with global formatting, [you can modify the header](http://rmarkdown.rstudio.com/html_document_format.html).

```{}
---
title: "Untitled"
author: "Name of a person"
date: "March 28, 2025"
output:
  html_document:
    toc: true # including a table of contents
    theme: readable # using a theme called readable
---
```

#### R Markdown Tinkering
R Markdown docs can be modified in many ways. Visit these links for more information.

* [Ways to modify the overall document appearance](http://rmarkdown.rstudio.com/html_document_format.html)
* [Ways to format parts of your document](http://rmarkdown.rstudio.com/authoring_basics.html)
* [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/)

<!-- TODO: Using Markdown for your Resume / CV-->

### 📚️ TODO:Using Markdown for your Resume / CV

**4 points**

*Create a brief Resume / CV using R Markdown with your name as level 4 headings, headings for (at least) education or employment. Each of the sections should include a bullet list of jobs/degrees. Highlight the year in bold. Try and incorporate as many R Markdown features as possible. You can work on this document or create a new markdown file.*

### R Code in R Markdown

Inside RMarkdown, lines of R code are called **chunks**. Code is sandwiched between sets of three backticks and `{r}`. This chunk of code...

    ```{r}`r ''`
    mean(c(1,2,3,4,5,6))
    ```

Produces this output in your document:
```{r}
mean(c(1,2,3,4,5,6)) # calculate the mean value of a list of numbers
```

Most commonly used two shortcuts for chunks:

| Action                   | Mac Shortcut              | Windows Shortcut         |
|--------------------------|---------------------------|--------------------------|
| Insert R code chunk      | ⌘ + Option + I            | Ctrl + Alt + I           |
| Run current chunk        | ⌘ + Option + C            | Ctrl + Alt + C           |

#### Chunk Options

Chunks have options that control what happens with their code, such as:

* `echo=FALSE`: Keeps R code from being shown in the document
* `eval=FALSE`: Shows R code in the document without running it
* `include=FALSE`: Hides all output but still runs code (good for `setup` chunks where you load packages!)
* `results='hide'`: Hides R's (non-plot) output from the document
* `cache=TRUE`: Saves results of running that chunk so if it takes a while, you won't have to re-run it each time you re-knit the document
* `fig.height=5, fig.width=5`: modify the dimensions of any plots that are generated in the chunk (units are in inches)

Some of these can be modified using the gear-shaped *Modify Chunk Options* button in each chunk. [There are a *lot* of other options, however](https://yihui.name/knitr/options/).


#### Editing and Running Code

In R Markdown documents, click within a code chunk and click the green arrow to run the chunk. The button beside that runs *all prior chunks*. The console will show the lines you ran followed by any printed output. There are more options on the top right of this window (green arrow with `Run`). 

```{r}
pi # output the value of pi
```

Sometimes we want to insert a value directly into our text. We do that using code in single backticks starting off with `r`. For example, the value of `pi` rounded to the nearest three decimals is `r round(pi, 3)`.

### Math and Logical Operation in R

```{r results='hide'}
# Math operations
2 + 3      # addition → 5
5 - 2      # subtraction → 3
4 * 6      # multiplication → 24
10 / 4     # division → 2.5
2^3        # exponentiation → 8
10 %% 3    # modulo (remainder) → 1
sqrt(400)  # square root of 400 → 20

x <- c(2, 4, 6, 8, 10) # Create those numbers to calculate

mean(x)      # average → 6
var(x)       # variance
sd(x)        # standard deviation
sum(x)       # sum of all elements
min(x)       # smallest value
max(x)       # largest value
range(x)     # min and max together
length(x)    # number of elements
```

```{r results='hide'}
# Logical operations
3 == 3     # equal → TRUE
3 != 4     # not equal → TRUE
5 > 2      # greater than → TRUE
2 < 1      # less than → FALSE
5 >= 5     # greater or equal → TRUE
2 <= 3     # less or equal → TRUE
TRUE & FALSE   # AND → FALSE
TRUE | FALSE   # OR → TRUE
!TRUE          # NOT → FALSE
```

#### Functions and Help

`sqrt()` is an example of a **function** in R. If we didn't have a good guess as to what `sqrt()` will do, we can type `?sqrt` in the console and look at the **Help** panel on the right.

```{r eval=FALSE}
?sqrt # what is sqrt()? 
```

**Arguments** are the *inputs* to a function. In this case, the only argument to `sqrt()`
is `x` which can be a number or a vector of numbers. *Help* files provide documentation on how to use functions and what functions produce.

#### Common Data Types in R

R has several *basic data types* that are used to build larger data structures.  

| Data Type   | Example                        | Description                                  | Check with |
|-------------|--------------------------------|----------------------------------------------|------------|
| **Numeric** | `x <- 3.142`                   | Decimal numbers (floating point)             | `is.numeric(x)` |
| **Integer** | `y <- 5L`                      | Whole numbers (the **L** forces integer type)  | `is.integer(y)` |
| **Character** | `z <- "Hello"`               | Text or string values                        | `is.character(z)` |
| **Logical** | `flag <- TRUE`                 | Boolean values: **TRUE** or **FALSE**            | `is.logical(flag)` |
| **Factor**  | `f <- factor(c("A", "B", "C"))`| Categorical values with fixed levels         | `is.factor(f)` |

You can also check type with `class()` or `is.*()` functions.

### Data Structures in R

#### Objects

R stores everything as an **object**, including data, functions, models, and output. Creating an object can be done using the assignment operator: `<-`: 

```{r}
new_object_preferred <- 144
new_object_not_preferred = 144 # using `=` is OK in R but not recommended; `=` is mainly for function arguments
```

The *Environment Pane* (right top) shows all active objects once you created in your current R session. Each assignment (`<-`) updates this list. Then. you can display or 'use' an object simply by using its name.

```{r}
new_object_preferred
sqrt(new_object_preferred)
```

Object names can contain `_` (preferred) and `.` in them but CANNOT begin with numbers. Try
to be consistent and clear in naming objects. Clear naming will help you save lots of time in the future!

#### Vectors

Vector is a sequence of elements of the same type.

```{r}
numbers <- c(1, 2, 3, 4)       # numeric vector; `c()` means combine
names <- c("Alice", "Bob")     # character vector
```

After creating them, we often want to inspect what they contain and their structure.

```{r}
print(numbers)           # show the values
length(numbers)          # check the number of elements
class(numbers)           # check the type
numbers > 2              # check whether each element is larger than 2
```

We may also want to select some elements in one vector. 

```{r}
numbers[1]            # show the first element
numbers[-1]           # show the vector without the first element
numbers[1:3]          # show the elements with index 1 to 3
numbers[c(1,3)]       # show the elements with index 1 and 3
numbers[c(TRUE, TRUE, FALSE)]       # show the elements with `TRUE`
```

#### Martix

A 2D collection of elements of the same type (optional for this class).

```{r}
mat <- matrix(1:6, nrow = 2, ncol = 3)
dim(mat)             # dimensions (# of rows, # of columns)
```

#### Lists

A collection of objects (can be of different types). List members can be named, referenced using the `[[` operator or using the names. *Note*: `list[ ]` will keep the result as a list (still wrapped).

```{r}
my_list <- list(name = "Alice", age = 25, scores = c(90, 85, 88))
```

```{r}
my_list$age     # get element by name ("age" = 25)
my_list[[2]]    # get element by position (2nd element = 25)
my_list[2]      # still a list with "age = 25"
```

#### Dataframes

Most data sets you will work with will be read into R and stored as a **dataframe**, so this course will mainly focus on manipulating and visualizing these objects. We will leave dataframes to next lab session.

<!-- TODO: R Basic Review Tasks-->

### 📚 TODO: R Basic Review Tasks

**5 points**

*What's your birthday (or any special day for you)? Please print it out using the format `YYYY-MM-DD` without showing the R code in the output HTML document. What's the data type of your `YYYY-MM-DD` object?*

```{r}
# TODO
```

*In R, a date can be defined as a class called `date` using the function `as.Date()`. Loop up the documentation of this function and tell me what's the simplest argument this function needs. Then, assign your birthday to an object.*

```{r}
# TODO
```

*Let's find the date for today! Sorry for the stupid request, but we need to use a function called `Sys.Date()` to always get "today". Check the documentation, what argument does we need?* 

```{r}
# TODO
```

*Create a vector with those two dates.*

```{r}
# TODO
```

*Let's use indexing `[]` to select today's date, then subtract your birthday. What's happened?*

```{r}
# TODO
```

*Can you calculate your age by simply dividing the difference in days by `365`? Note: The calculation doesn’t account for the exact month/day alignment — it just assumes each year is a flat 365 days. You can install package `lubridate` to calculate a more precise age (optional).*

```{r}
# TODO
```

*Then, finish the sentence using inline R code:*

Today is `today's date`, it has been `number of year` years since {what did happen on the big day, such as my birthday}. 

---

## Lab 01-C: Dataframes and Accessing Census Data

### Working Directory

```{r}
# show the current working directory
getwd()
# TODO: modify the folder stored this `.Rmd` as the working directory
setwd("/Users/yohaoyu/Repo/data-analytics-visualization/labs") 
```

- `.Rmd` files use their current directory as a working directory.
- For larger projects, instead of setting a working directory, it is usually better to use [RStudio projects](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects) to manage working directories.
- Windows users: If you copy a path from Explorer, make sure to change back slashes (`\`) to forward slashes (`/`) for the file paths

### Access Census Data

In this lab, we will explore datasets from [American Community Survey (ACS)](https://www.census.gov/programs-surveys/acs.html) via [DATA.CENSUS.GOV](https://data.census.gov) and R package [`tidycensus`](https://walker-data.com/tidycensus/). 

We will use 2023 ACS 5-year estimate for B25075 Value, which means the estimated selling prices for occupied housing units, as an example. Some intro to this data from Census Reporter: <https://censusreporter.org/tables/B25075/>.

To use `tidycensus`, we need load the package and have a Census API key. A key can be obtained from <http://api.census.gov/data/key_signup.html>.

```{r message=FALSE}
# remove the following '#' to install the `tidycensus` and `tidyverse` package; `tidyverse` will be most common-used package for use in the future
# install.packages("tidycensus")
# install.packages("tidyverse")

# This loads packages into the current session; however, it is best practice to load all packages at the start of an R Markdown document
library(tidycensus) 
library(tidyverse)
```

```{r}
census_api_key("96f943fd5b38142362574386830d3c855e82dd28")
```

```{r}
king_value <- get_acs(geography = "tract", # extract all tract-level data
            table = "B25031", 
            state = "53", # 53 means Washington State, you can use 'WA'
            county = "033", # 033 means King County, you can use 'king'
            year = 2023)
```

We can save this `king_value` tibble (just like a table) on our local machine as a `.csv` file. `.csv` is one of the most common-used file types in the world of data science. But, [what is a CSV file?](https://www.adobe.com/acrobat/resources/document-files/what-is-a-csv-file.html)

```{r}
# The CSV file will be saved in your current working directory
# Here, we use the function `write_csv` from package `readr`, which is built within `tidyverse`
write_csv(king_value, "king_value.csv") 
```

Once you've set the working directory—or you're in an RStudio project—you can refer to folders and files within the working directory using relative paths, like we've done in previous codes - saving the file in the current working directory. Then, we can also import this `.csv` again using `read_csv()`:

```{r}
another_king_value <- read_csv("king_value.csv")
```

### Dataframes

Most data sets you will work with will be read into R and stored as a **dataframe**, just like the `king_value` data. So this course will mainly focus on manipulating and visualizing these objects. 

```{r}
# check whether an object is a dataframe
is.data.frame(king_value)
```

You can check the dataframe by clicking the object name in the *Environment Panel* (right top) or using `view()` function.

```{r}
# check the dataframe
view(king_value)

# check the first three rows, default as five rows
head(king_value, 3)

# check the column names
colnames(king_value)

# number of rows and columns
dim(king_value)

# select GEOID column and elements with index 1 to 10
king_value$GEOID[1:10]

# calculate the average estimate
# na.rm = TRUE helps to ignore missing values
mean(king_value$estimate, na.rm = TRUE)
```

When you click the **Knit** button (top of this panel) a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. 


<!-- TODO: Explore Census Demographic Data -->

### 📚 TODO: Explore Census Demographic Data

**7 points**

*Let's do the similar step for other variables, select 1~2 interesting datasets from ACS, download them via API or via DATA.CENSUS.GOV. You may look up variables you're interested in from [this Census Reporter site](https://censusreporter.org/topics/table-codes/) or [ACS Technical Documentation](https://www.census.gov/programs-surveys/acs/technical-documentation.html/).*

*Using a dataframe to store the data and calculate basic summary statistics such as mean, median, min/max, 25th/75th percentiles for your data; and briefly introduce why you're interested in this data and what did you find through the simple exploration.*

*After finish this lab, you need to submit everything via Canvas assignment page.*

```{r}
# TODO
```

## Acknowledgement

The materials are developed by [Haoyu Yue](www.yuehaoyu.com) based materials from [Dr. Feiyang Sun at UC San Diego](https://fsun.ucsd.edu), Siman Ning and Christian Phillips at University of Washington, [Dr. Charles Lanfear at University of Cambridge](https://clanfear.github.io).
